Nondeterministic Bigraphical Reactive Systems for Markov Decision Processes\thanks{Supported by organization x.}

### 2018-08-15 paper.tex
And another example

\begin{example}
  Let $M = (S, \overline{s}, A, P, L)$ be an MDP, where $S = \{ s_0, s_1, s_2,
  s_3 \}$, $\overline{s} = s_0$, $A = \{ \text{start}, \text{wait}, \text{send},
  \text{restart}, \text{stop} \}$, and $P, L$ are defined as follows:
  \begin{equation*}
    \begin{split}
      P(s_0, \text{start}) &= P(s_1, \text{wait}) = [s_1 \mapsto 1], \\
      P(s_1, \text{send}) &= [s_2 \mapsto 0.01, s_3 \mapsto 0.99], \\
      P(s_2, \text{restart}) &= [s_0 \mapsto 1], \\
      P(s_3, \text{stop}) &= [s_3 \mapsto 1],
    \end{split}
    \qquad
    \begin{split}
      L(s_0) &= \emptyset, \\
      L(s_1) &= \{ \text{try} \}, \\
      L(s_2) &= \{ \text{failure} \}, \\
      L(s_3) &= \{ \text{success} \}.
    \end{split}
  \end{equation*}
  Then $M$ can represent a simple communication protocol. The process
  initialises, and is faced with a nondeterministic choice between waiting and
  sending a message. Most of the time the sent message is successfully
  delivered, and the process stops. But sometimes a message fails to deliver, in
  which case the process has to restart. See Fig. \ref{first_mdp} for a
  visualisation of $M$.
\end{example}

@@

\begin{example}
  Consider a simple PBRS, where the initial state $g_0$ has an \texttt{Agent}
  and two \texttt{Locations}, each of which is holding an \texttt{Object} (see
  Fig. \ref{pbrs_initial}), and a single reaction rule with probability $1$ that
  takes an \texttt{Object} from a \texttt{Location} and transfers it to the
  \texttt{Agent} (see Fig. \ref{pbrs_rule}). Then, since regions $1$ and $2$ are
  not treated as isomorphic, the reaction rule can be applied in two ways,
  resulting in two different subsequent states in Fig. \ref{pbrs_next_states},
  $g_1$ and $g_2$. Thus, we have two ground probabilistic reaction rules ($n =
  2$) with $p_1 = p_2 = 1$ and $s = 2$. Let $g' = g_1$. Then $|S| = 1$, and
  \[ g_0 \xrightarrowrhd{0.5}_{\mathcal{R}} g_1, \]
  since
  \[ 0.5 = \frac{1}{2} \sum_{(r, r', 1) \in S} 1. \]
  Similarly,
  \[ g_0 \xrightarrowrhd{0.5}_{\mathcal{R}} g_2. \]
\end{example}

@@

  Let $g, g'$ be ground bigraphs, $a \in A$ an action, and $\{ (r_i, r'_i, a,
  p_i) \}_{i=1}^n$ a set of ground nondeterministic reaction rules with action
  $a$, where for each $r_i$, there exists a bigraph $D(r_i)$ such that $g =
  D(r_i) \circ r_i$. Let $S = \{ (r_i, r'_i, a, p_i) \mid g' = D(r_i) \circ r'_i
  \}$, and
  \[ s = \sum_{i=1}^n p_i. \]
  Then the reaction relation for action $a$ is defined as
  \[ g \xrightarrowrhd[\rho(a)]{p}_{a} g' \text{ iff } S \ne \emptyset, \]
  where
  \[ p = \frac{1}{s}\sum_{(r, r', a, p') \in S} p'. \]
\end{definition}

@@

TODO: minimal example, possibly the same one as below.

@@

The described extensions have been implemented in an open-source tool for BRSs
BigraphER \cite{DBLP:conf/cav/SevegnaniC16}. We refer the interested reader to
the paper on BigraphER \cite{DBLP:conf/cav/SevegnaniC16}, the related PhD thesis
\cite{DBLP:phd/ethos/Sevegnani12}, and online
resources\footnote{\url{http://www.dcs.gla.ac.uk/~michele/bigrapher.html}} for
further information on the syntax and capabilities of the software. In this
section we will introduce the new features of BigraphER by using a simple MDP as
an example.

\begin{example} \label{example}
  Consider an MDP $(S, \overline{s}, A, P, L)$, where $S = \{ s_0, s_1, s_2, s_3
  \}$, $\overline{s} = s_0$, $A = \{ a, b, c \}$, and $P, L$ are defined as
  follows:
  \begin{equation*}
    \begin{split}
      P(s_0, a) &= [s_1 \mapsto 1], \\
      P(s_1, b) &= [s_0 \mapsto 0.7, s_1 \mapsto 0.3], \\
      P(s_1, c) &= [s_2 \mapsto 0.5, s_3 \mapsto 0.5], \\
      P(s_2, a) &= [s_2 \mapsto 1], \\
      P(s_3, a) &= [s_3 \mapsto 1],
    \end{split}
    \qquad
    \begin{split}
      L(s_0) &= \{ \text{initial} \}, \\
      L(s_1) &= \emptyset, \\
      L(s_2) &= \{ \text{heads} \}, \\
      L(s_3) &= \{ \text{tails} \}.
    \end{split}
  \end{equation*}
  Furthermore, equip it with a reward structure $(\rho, \iota)$, where
  $\rho(s_2) = 3$, $\iota(s_1, b) = 1$, and both functions are zero everywhere
  else.
\end{example}

@@

The MDP can be represented as an NBRS with BigraphER code in
Listing~\ref{example_listing}. More specifically, reaction rule probabilities
are represented as floating-point numbers inside the arrows (e.g.
\texttt{-[0.7]->}), each action encompasses its reaction rules with
\texttt{\textbf{action} actionName} and \texttt{\textbf{end}}. Rewards can be
added to actions simply by inserting an integer enclosed in square brackets
after the action name (e.g. \texttt{\textbf{action} b[1]}). Lastly, predicate
rewards have the same format, but are listed on the \texttt{\textbf{preds}} line
of the \texttt{\textbf{begin nbrs}}---\texttt{\textbf{end}} block (e.g.
\texttt{\textbf{preds} = \{heads[3]\};}).

@@

We introduce a new front end to BigraphER based on Jupyter notebooks
\cite{DBLP:conf/elpub/KluyverRPGBFKHG16}. The BigraphER kernel is an extension
to the OCaml kernel\footnote{\url{https://github.com/akabe/ocaml-jupyter}} and
is available on
GitHub\footnote{\url{https://github.com/dilkas/bigrapher-jupyter}}. It supports
a similar workflow to that of Jupyter notebooks for other languages, allowing
the user to divide the code into multiple \emph{cells} and have the definitions
persist in a \emph{buffer} to be used later. By default, the output of each cell
displays the graphical representation of all bigraphs and reaction rules defined
in that cell (see Fig. \ref{interface} for an example). The generated images are
stored in a local directory \texttt{jupyter-images/}. The kernel also supports a
number of \emph{magics}, i.e., custom commands that can be used at the top of a
cell:

\begin{description}
  \item[\texttt{\%states}] produces a state transition diagram (similar to the
    one in Fig. \ref{example_ts}). Mousing over a state shows what that
    state looks like as a bigraph.
  \item[\texttt{\%simulate n}] runs a simulation for \texttt{n} steps, when used
    on non-stochastic models. For stochastic BRSs
    \cite{DBLP:journals/entcs/KrivineMT08}, \texttt{n} is interpreted as a
    floating-point number, representing maximum simulation time. The produced
    diagram is a single path through the transition system with the same
    mouse-over functionality.
  \item[\texttt{\%output}] displays the full output produced by BigraphER. By
    default, only errors are displayed in the notebook.
  \item[\texttt{\%ocaml}] allows full backward compatibility with the OCaml
    kernel, including auto-complete and integrated documentation features that
    have been extended to the BigraphER OCaml API as well as plotting support
    with Archimedes\footnote{\url{http://archimedes.forge.ocamlcore.org/}}.
  \item[\texttt{\%clear}] clears the buffer before interpreting the code from
    the current cell.
\end{description}

@@

In order to implement nesting, we put a \texttt{Node} (representing a room)
inside a \texttt{Cell}, which has the ``door'' to that room (see Fig.
\ref{agent2_nesting}). The \texttt{Node} can have its own collection of
\texttt{Cell}s, describing the topology of the room. One of these \texttt{Cell}s
is linked to the \texttt{Node}, representing the \texttt{Cell} closest to the
door from the inside of the room. Thus, both \texttt{Node}s and \texttt{Cell}s
have arity $1$ for exactly this linking. \texttt{Cell} ports not linked to
anything are implemented using closures (e.g. \texttt{/x Cell\{x\}}). The
initial map in Listing~\ref{agent2_initial} has a corridor of three cells with
the agent on the west-most cell and a door to a room in the middle cell. One
moves in and out of a room using a pair of actions with reaction rules pictured
in Fig.~\ref{go_in_out}.

@@

The room is a $2 \times 2$ grid of \texttt{Cell}s, which is generated when the
\texttt{Agent} gets to a \texttt{Cell} with an empty \texttt{Node} (represented
by \texttt{Node\{n\}.1})\footnote{Multiple different rooms can be implemented by
having each previously empty \texttt{Node} contain a node of different control.
This way a different probability distribution (with different possible outcomes)
can be assigned to each room.}. In order for the room to be generated before any
other action can take place, the two reaction rules in Listing~\ref{agent2_room}
are put on a higher priority class. With probability $p$, the generated room has
each \texttt{Cell} linked to its two neighbours; and with probability $1-p$, the
top two \texttt{Cell}s are not linked, possibly forcing the agent to choose a
different path (see Fig. \ref{agent2_map}). Finally, we have four actions, each
with a single reaction rule with probability $1$, corresponding to the four
possible directions of travel. The reaction rules are very similar to those of
our previous example, an example of which can be found in
Listing~\ref{agent2_north}.

>>>>

\begin{example}
  Let $M = (S, \overline{s}, A, P, L)$ be an MDP, where $S = \{ s_0, s_1, s_2,
  s_3 \}$, $\overline{s} = s_0$, $A = \{ \text{start}, \text{wait}, \text{send},
  \text{restart}, \text{stop} \}$, and $P, L$ are defined as follows:
  \begin{equation*}
    \begin{split}
      P(s_0, \text{start}) &= P(s_1, \text{wait}) = [s_1 \mapsto 1], \\
      P(s_1, \text{send}) &= [s_2 \mapsto 0.01, s_3 \mapsto 0.99], \\
      P(s_2, \text{restart}) &= [s_0 \mapsto 1], \\
      P(s_3, \text{stop}) &= [s_3 \mapsto 1],
    \end{split}
    \qquad
    \begin{split}
      L(s_0) &= \emptyset, \\
      L(s_1) &= \{ \text{try} \}, \\
      L(s_2) &= \{ \text{failure} \}, \\
      L(s_3) &= \{ \text{success} \}.
    \end{split}
  \end{equation*}
  Then $M$ can represent a simple communication protocol. The process
  initialises, and is faced with a nondeterministic choice between waiting and
  sending a message. Most of the time the sent message is successfully
  delivered, and the process stops. But sometimes a message fails to deliver, in
  which case the process has to restart. See Fig.~\ref{first_mdp} for a
  visualisation of $M$.
\end{example}

@@

\begin{example}
  Consider a simple PBRS, where the initial state $g_0$ has an \texttt{Agent}
  and two \texttt{Locations}, each of which is holding an \texttt{Object} (see
  Fig.~\ref{pbrs_initial}), and a single reaction rule with probability $1$ that
  takes an \texttt{Object} from a \texttt{Location} and transfers it to the
  \texttt{Agent} (see Fig.~\ref{pbrs_rule}). Then, since regions $1$ and $2$ are
  not treated as isomorphic, the reaction rule can be applied in two ways,
  resulting in two different subsequent states in Fig.~\ref{pbrs_next_states},
  $g_1$ and $g_2$. Thus, we have two ground probabilistic reaction rules ($n =
  2$) with $p_1 = p_2 = 1$ and $s = 2$. Let $g' = g_1$. Then $|S| = 1$, and
  \[ g_0 \xrightarrowrhd{0.5}_{\mathcal{R}} g_1, \]
  since
  \[ 0.5 = \frac{1}{2} \sum_{(r, r', 1) \in S} 1. \]
  Similarly,
  \[ g_0 \xrightarrowrhd{0.5}_{\mathcal{R}} g_2. \]
\end{example}

@@

  Let $g, g'$ be ground bigraphs, $a \in A$ an action, and $\{ (r_i, r'_i, a,
  p_i) \}_{i=1}^n$ a set of ground nondeterministic reaction rules with action
  $a$, where for each $r_i$, there exists a bigraph $D(r_i)$ such that $g =
  D(r_i) \circ r_i$. Let $S = \{ (r_i, r'_i, a, p_i) \mid g' = D(r_i) \circ r'_i
  \}$, and
  \[ s = \sum_{i=1}^n p_i. \]
  Then the reaction relation for action $a$ is defined as
  \[ g \xrightarrowrhd[\rho(a)]{p}_a g' \text{ iff } S \ne \emptyset, \]
  where
  \[ p = \frac{1}{s}\sum_{(r, r', a, p') \in S} p'. \]
\end{definition}

@@

\begin{example} \label{example}
  Consider an MDP $(S, \overline{s}, A, P, L)$, where $S = \{ s_0, s_1, s_2,
  s_3 \}$, $\overline{s} = s_0$, $A = \{ a, b, c \}$, and $P, L$ are defined as
  follows:
  \begin{equation*}
    \begin{split}
      P(s_0, a) &= [s_1 \mapsto 1], \\
      P(s_1, b) &= [s_0 \mapsto 0.7, s_1 \mapsto 0.3], \\
      P(s_1, c) &= [s_2 \mapsto 0.5, s_3 \mapsto 0.5], \\
      P(s_2, a) &= [s_2 \mapsto 1], \\
      P(s_3, a) &= [s_3 \mapsto 1],
    \end{split}
    \qquad
    \begin{split}
      L(s_0) &= \{ \text{initial} \}, \\
      L(s_1) &= \emptyset, \\
      L(s_2) &= \{ \text{heads} \}, \\
      L(s_3) &= \{ \text{tails} \}.
    \end{split}
  \end{equation*}
  Furthermore, equip it with a reward structure $(\rho, \iota)$, where
  $\rho(s_2) = 3$, $\iota(s_1, b) = 1$, and both functions are zero everywhere
  else.

  Let $s_0, s_1, s_2, s_3$ also denote the bigraphs representing these states.
  Then we have the following reaction relations:
  \begin{align*}
    s_0 &\xrightarrowrhd{1}_a s_1, & s_1 &\xrightarrowrhd[1]{0.7}_b s_0, & s_1 &\xrightarrowrhd[1]{0.3}_b s_1, & s_1 &\xrightarrowrhd{0.5}_c s_2, & \\
    s_1 &\xrightarrowrhd{0.5}_c s_3, & s_2 &\xrightarrowrhd{1}_a s_2, & s_3 &\xrightarrowrhd{1}_a s_3,
  \end{align*}
  where zero rewards have been omitted for clarity.
\end{example}

@@

The described extensions have been implemented in an open-source tool for BRSs
BigraphER \cite{DBLP:conf/cav/SevegnaniC16}. We refer the interested reader to
the paper on BigraphER \cite{DBLP:conf/cav/SevegnaniC16}, the related PhD thesis
\cite{DBLP:phd/ethos/Sevegnani12}, and online
resources\footnote{\url{http://www.dcs.gla.ac.uk/~michele/bigrapher.html}} for
further information on the syntax and capabilities of the software. In this
section we will introduce the new features of BigraphER by turning Example
\ref{example} into an NBRS.

@@

The full BigraphER code for this example is in Listing~\ref{example_listing}.
Reaction rule probabilities are represented as floating-point numbers inside the
arrows (e.g. \texttt{-[0.7]->}), each action encompasses its reaction rules with
\texttt{\textbf{action} actionName} and \texttt{\textbf{end}}. Rewards can be
added to actions simply by inserting an integer enclosed in square brackets
after the action name (e.g. \texttt{\textbf{action} b[1]}). Lastly, predicate
rewards have the same format, but are listed on the \texttt{\textbf{preds}} line
of the \texttt{\textbf{begin nbrs}}---\texttt{\textbf{end}} block (e.g.
\texttt{\textbf{preds} = \{heads[3]\};}).

@@

We introduce a new front end to BigraphER based on Jupyter notebooks
\cite{DBLP:conf/elpub/KluyverRPGBFKHG16}. The BigraphER kernel is an extension
to the OCaml kernel\footnote{\url{https://github.com/akabe/ocaml-jupyter}} and
is available on
GitHub\footnote{\url{https://github.com/dilkas/bigrapher-jupyter}}. It supports
a similar workflow to that of Jupyter notebooks for other languages, allowing
the user to divide the code into multiple \emph{cells} and have the definitions
persist in a \emph{buffer} to be used later. By default, the output of each cell
displays the graphical representation of all bigraphs and reaction rules defined
in that cell (see Fig.~\ref{interface} for an example). The generated images are
stored in a local directory \texttt{jupyter-images/}. The kernel also supports a
number of \emph{magics}, i.e., custom commands that can be used at the top of a
cell:

\begin{description}
  \item[\texttt{\%states}] produces a state transition diagram (similar to the
    one in Fig.~\ref{example_ts}). Mousing over a state shows what that
    state looks like as a bigraph.
  \item[\texttt{\%simulate n}] runs a simulation for \texttt{n} steps, when used
    on non-stochastic models. For stochastic BRSs
    \cite{DBLP:journals/entcs/KrivineMT08}, \texttt{n} is interpreted as a
    floating-point number, representing maximum simulation time. The produced
    diagram is a single path through the transition system with the same
    mouse-over functionality.
  \item[\texttt{\%output}] displays the full output produced by BigraphER. By
    default, only errors are displayed in the notebook.
  \item[\texttt{\%ocaml}] allows full backward compatibility with the OCaml
    kernel, including auto-complete and integrated documentation features that
    have been extended to the BigraphER OCaml API as well as plotting support
    with Archimedes\footnote{\url{http://archimedes.forge.ocamlcore.org/}}.
  \item[\texttt{\%clear}] clears the buffer before interpreting the code from
    the current cell.
\end{description}

@@

In order to implement nesting, we put a \texttt{Node} (representing a room)
inside a \texttt{Cell}, which has the ``door'' to that room (see
Fig.~\ref{agent2_nesting}). The \texttt{Node} can have its own collection of
\texttt{Cell}s, describing the topology of the room. One of these \texttt{Cell}s
is linked to the \texttt{Node}, representing the \texttt{Cell} closest to the
door from the inside of the room. Thus, both \texttt{Node}s and \texttt{Cell}s
have arity $1$ for exactly this linking. \texttt{Cell} ports not linked to
anything are implemented using closures (e.g. \texttt{/x Cell\{x\}}). The
initial map in Listing~\ref{agent2_initial} has a corridor of three cells with
the agent on the west-most cell and a door to a room in the middle cell. One
moves in and out of a room using a pair of actions with reaction rules pictured
in Fig.~\ref{go_in_out}.

@@

The room is a $2 \times 2$ grid of \texttt{Cell}s, which is generated when the
\texttt{Agent} gets to a \texttt{Cell} with an empty \texttt{Node} (represented
by \texttt{Node\{n\}.1})\footnote{Multiple different rooms can be implemented by
having each previously empty \texttt{Node} contain a node of different control.
This way a different probability distribution (with different possible outcomes)
can be assigned to each room.}. In order for the room to be generated before any
other action can take place, the two reaction rules in Listing~\ref{agent2_room}
are put on a higher priority class. With probability $p$, the generated room has
each \texttt{Cell} linked to its two neighbours; and with probability $1-p$, the
top two \texttt{Cell}s are not linked, possibly forcing the agent to choose a
different path (see Fig.~\ref{agent2_map}). Finally, we have four actions, each
with a single reaction rule with probability $1$, corresponding to the four
possible directions of travel. The reaction rules are very similar to those of
our previous example, an example of which can be found in
Listing~\ref{agent2_north}.
